---
title: "Amplicons AMOVA analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Init environment and import packages
```{r setup, include=FALSE}
setwd("~/Desktop/CRISPR_distance/manuscript_data_and_downstream_analyses_scripts")
knitr::opts_chunk$set(echo = TRUE)
require('dplyr')
require('reshape2')
library('ggplot2')
require('ade4')
require('usedist') # package used to subset distance matrix
require('poppr')
require('vegan')
require('ape')
library(stringr)
```

# Init functions

```{r}
fix_phi_df <- function(df){
  df1 <- arrange(df, -row_number())
  rownames(df1) <- rev(rownames(df))
  df1=rbind(df1,data.frame(row.names = 'Total',Phi = NA))
  return(df1)
}

build_results_table <- function(res) {
  
  pval = randtest(res, nrepet =999)
  
  pval = rbind(data.frame(P.value=rev(pval$pvalue),row.names = rev(pval$names)),data.frame(row.names = 'Total',P.value = NA))
  
  table=cbind(res$results,res$componentsofcovariance,pval,fix_phi_df(res$statphi))
  
  return(table)
  
}



```


# Import data. One haplotype per sequence read
## Import Amplicons data
```{r}

#### Import Amplicons allele counts in populations ####
path_to_Amplicons_allele_counts_in_pop = './Amplicons_haplotypes_all.txt'

Amplicons_all = read.table(path_to_Amplicons_allele_counts_in_pop,header=TRUE,sep='\t')

colnames(Amplicons_all) = c('Sample', colnames(Amplicons_all) %>% gsub('\\.','-',.) %>% .[2:length(.)])
Amplicons_all[is.na(Amplicons_all)] <- 0
rownames(Amplicons_all) = as.character(Amplicons_all$Sample) %>% substr(., 13, nchar(.))

Amplicons_all %>% head


### Import metdata ###
path_to_Amplicons_metadata = '~/Dropbox/CRISPR_haplo/Amplicons_metadata.txt'

Amplicons_meta=read.table(path_to_Amplicons_metadata,header=T,sep='\t',as.is = c("Site",'Region','Flow','Ind','Duplicate'))
Amplicons_meta$Sample = as.character(Amplicons_meta$Sample) %>% substr(., 5, nchar(.))
row.names(Amplicons_meta)=Amplicons_meta$Sample
colnames(Amplicons_meta)[7]='dupl'
Amplicons_meta$chunk = substr(Amplicons_meta$Sample, nchar(Amplicons_meta$Sample)-1, nchar(Amplicons_meta$Sample))
Amplicons_meta$chunk[Amplicons_meta$chunk %in% c('-T','is','ld')] = 'full'

Amplicons_meta=Amplicons_meta[,-c(1,2,8)]
Amplicons_meta %>% head(.)


Amplicons_haplos=merge(Amplicons_all,Amplicons_meta, by=0,all.x=T)
rownames(Amplicons_haplos)=Amplicons_haplos$Row.names
Amplicons_haplos=Amplicons_haplos[,3:ncol(Amplicons_haplos)]

Amplicons_haplos %>% head()
str(Amplicons_haplos)

#### Import per-read allele counts in populations. #### 

path_to_LpxA_per_read_allele_counts_in_pop = './LpxA_per_read.txt'
path_to_PleD_per_read_allele_counts_in_pop = './PleD_per_read.txt'

Amplicons_LpxA_per_read <- read.table(path_to_LpxA_per_read_allele_counts_in_pop,header=T,row.names = 1,sep='\t')
Amplicons_LpxA_per_read

Amplicons_PleD_per_read <- read.table(path_to_PleD_per_read_allele_counts_in_pop,header=T,row.names = 1,sep='\t')
Amplicons_PleD_per_read

#### Import Phi-st distance matrix #### 

path_to_LpxA_Phist_matrix = './Amplicons_LpxA_pairwiseFst.txt'
path_to_LpxA_Phist_pval_matrix = './Amplicons_LpxA_pairwiseFst_pval.txt'

path_to_PleD_Phist_matrix = './Amplicons_PleD_pairwiseFst.txt'
path_to_PleD_Phist_pval_matrix = './Amplicons_PleD_pairwiseFst_pval.txt'

# load phi-st matrices
LpxA_phist=read.csv(path_to_LpxA_Phist_matrix,header=T,row.names = 1,sep='\t')
colnames(LpxA_phist) = c(colnames(LpxA_phist) %>% gsub('\\.','-',.) %>% .[1:length(.)])
LpxA_phist %>% head

PleD_phist=read.csv(path_to_PleD_Phist_matrix,header=T,row.names = 1,sep='\t')
colnames(PleD_phist) = c(colnames(PleD_phist) %>% gsub('\\.','-',.) %>% .[1:length(.)])
PleD_phist %>% head


# load pvalue matrices and turned then into filter matrices whose role will be to remove instances of phi-st where max_pval > 0.05
LpxA_phist_pval=read.csv(path_to_LpxA_Phist_pval_matrix,header=T,row.names = 1,sep='\t',as.is = T)
colnames(LpxA_phist_pval) = c(colnames(LpxA_phist_pval) %>% gsub('\\.','-',.) %>% .[1:length(.)])

LpxA_phist_pval[LpxA_phist_pval == "*"] <- 0
LpxA_phist_pval[LpxA_phist_pval == ""] <- NA
LpxA_phist_pval = LpxA_phist_pval %>% mutate_all(funs(str_replace_all(., "\\+-", "|")))
rownames(LpxA_phist_pval) = colnames(LpxA_phist_pval)
LpxA_phist_pvalmax = LpxA_phist_pval %>% lapply(.,function(x) sapply(sapply((strsplit(x,"\\|")),as.numeric),sum)) %>% as.data.frame
colnames(LpxA_phist_pvalmax) = c(colnames(LpxA_phist_pval))
rownames(LpxA_phist_pvalmax) = c(colnames(LpxA_phist_pval))
LpxA_phist_pvalmax[LpxA_phist_pvalmax > 0.05] <- NA
LpxA_phist_pvalmax[LpxA_phist_pvalmax < 0.05] <- 1
LpxA_phist_pvalmax %>% head


PleD_phist_pval=read.csv(path_to_PleD_Phist_pval_matrix,header=T,row.names = 1,sep='\t',as.is = T)
colnames(PleD_phist_pval) = c(colnames(PleD_phist_pval) %>% gsub('\\.','-',.) %>% .[1:length(.)])

PleD_phist_pval[PleD_phist_pval == "*"] <- 0
PleD_phist_pval[PleD_phist_pval == ""] <- NA
PleD_phist_pval = PleD_phist_pval %>% mutate_all(funs(str_replace_all(., "\\+-", "|")))
rownames(PleD_phist_pval) = colnames(PleD_phist_pval)
PleD_phist_pvalmax = PleD_phist_pval %>% lapply(.,function(x) sapply(sapply((strsplit(x,"\\|")),as.numeric),sum)) %>% as.data.frame
colnames(PleD_phist_pvalmax) = c(colnames(PleD_phist_pval))
rownames(PleD_phist_pvalmax) = c(colnames(PleD_phist_pval))
PleD_phist_pvalmax[PleD_phist_pvalmax > 0.05] <- NA
PleD_phist_pvalmax[PleD_phist_pvalmax < 0.05] <- 1
PleD_phist_pvalmax %>% head

# filter phi-st tables based on pval and sort the distance matrices in alphabetical order

filtered_LpxA_phist = LpxA_phist * LpxA_phist_pvalmax 
filtered_LpxA_phist = filtered_LpxA_phist %>% as.matrix() %>% as.dist()
LpxA_phist = dist_subset(filtered_LpxA_phist,idx=sort(labels(filtered_LpxA_phist)))
LpxA_phist
LpxA_phist_nomv=dist_subset(LpxA_phist,idx = END_samples)

filtered_PleD_phist = PleD_phist * PleD_phist_pvalmax 
filtered_PleD_phist = filtered_PleD_phist %>% as.matrix() %>% as.dist()
PleD_phist = dist_subset(filtered_PleD_phist,idx=sort(labels(filtered_PleD_phist)))
PleD_phist
PleD_phist_nomv=dist_subset(PleD_phist,idx = END_samples)


```


# Duplicates 
Is the distribution of Amplicons haplotypes the same between technical duplicates? 
Significant differences (Pval <0.1) were found for LpxA R13L-5-T and R16L-3-T

```{r}
#### Amplicons ####
LpxA_dupl1=Amplicons_all[c("R13L-5-T","R13L-5-T_bis"),2:5]
LpxA_dupl1_nozero=LpxA_dupl1[rowSums(LpxA_dupl1)>0,]
fisher.test(LpxA_dupl1)
diversity(t(LpxA_dupl1))  %>% mean(.)

LpxA_dupl2=Amplicons_all[c("R15H-3-T","R15H-3-T_old"),2:5]
LpxA_dupl2_nozero=LpxA_dupl2[rowSums(LpxA_dupl2)>0,]
fisher.test(LpxA_dupl2)
diversity(t(LpxA_dupl2))  %>% mean(.)

LpxA_dupl3=Amplicons_all[c("R16L-3-T","R16L-3-T_bis"),2:5]
LpxA_dupl3_nozero=LpxA_dupl3[rowSums(LpxA_dupl3)>0,]
fisher.test(LpxA_dupl3)
diversity(t(LpxA_dupl3))  %>% mean(.)


PleD_dupl1=Amplicons_all[c("R13L-5-T","R13L-5-T_bis"),6:9]
PleD_dupl1_nozero=PleD_dupl1[rowSums(PleD_dupl1)>0,]
fisher.test(PleD_dupl1)
diversity(t(PleD_dupl1))  %>% mean(.)

PleD_dupl2=Amplicons_all[c("R15H-3-T","R15H-3-T_old"),6:9]
PleD_dupl2_nozero=PleD_dupl2[rowSums(PleD_dupl2)>0,]
fisher.test(PleD_dupl2)
diversity(t(PleD_dupl2))  %>% mean(.)

PleD_dupl3=Amplicons_all[c("R16L-3-T","R16L-3-T_bis"),6:9]
PleD_dupl3_nozero=PleD_dupl3[rowSums(PleD_dupl3)>0,]
fisher.test(PleD_dupl3)
diversity(t(PleD_dupl3))  %>% mean(.)

```


# Run ade4's AMOVA analysis using the Poppr wrapper

## Create genind object

Add info about identical technical replicates to meta data
```{r}

Amplicons_meta$DedupInd<-ifelse(rownames(Amplicons_meta) == 'R13L-5-T_bis','R13L-5-T_bis',
                      ifelse(rownames(Amplicons_meta) == 'R16L-3-T_bis','R16L-3-T_bis', Amplicons_meta$Ind))
Amplicons_meta$Sample = rownames(Amplicons_meta)

head(Amplicons_meta)
str(Amplicons_meta)
```


```{r}


LpxA_per_read_meta = merge(Amplicons_LpxA_per_read[,c('Sample','read_id.1')],Amplicons_meta[,c('Sample','Site','Ind','Region','Flow','dupl','chunk','DedupInd')], by='Sample', all.x=T)
rownames(LpxA_per_read_meta) = LpxA_per_read_meta$read_id.1
LpxA_per_read_meta$read_id.1 <- NULL
LpxA_per_read_meta %>% head

PleD_per_read_meta = merge(Amplicons_PleD_per_read[,c('Sample','read_id.1')],Amplicons_meta[,c('Sample','Site','Ind','Region','Flow','dupl','chunk','DedupInd')], by='Sample', all.x=T)
rownames(PleD_per_read_meta) = PleD_per_read_meta$read_id.1
PleD_per_read_meta$read_id.1 <- NULL
PleD_per_read_meta %>% head

LpxA=as.data.frame(Amplicons_LpxA_per_read[,'haplo'],row.names = row.names(Amplicons_LpxA_per_read))
colnames(LpxA)[1]='haplo'
LpxA %>% head

LpxA_obj <- df2genind(LpxA, ploidy=1,pop=factor(LpxA_per_read_meta$Sample), strata=LpxA_per_read_meta)

PleD=as.data.frame(Amplicons_PleD_per_read[,'haplo'],row.names = row.names(Amplicons_PleD_per_read))
colnames(PleD)[1]='haplo'
PleD %>% head

PleD_obj <- df2genind(PleD, ploidy=1,pop=factor(PleD_per_read_meta$Sample), strata=PleD_per_read_meta)

popsub(PleD_obj)
as.genclone(PleD_obj)
```

## AMOVA in the MEF (sites within Flow + trophosome sections within hosts)
```{r}
CB_MV=c('R13','R14','R15','R16')

MEF_samples = popNames(LpxA_obj)[!(popNames(LpxA_obj) %>% grepl(paste(CB_MV, collapse="|"),.))]

LpxA.MEF = popsub(LpxA_obj, sublist=MEF_samples)
PleD.MEF = popsub(PleD_obj, sublist=MEF_samples)

LpxA.MEF.amova = poppr.amova(LpxA.MEF,hier=~Flow/Ind/Sample,clonecorrect = FALSE,within=F,threads=4,method='ade4')
# randtest(LpxA.MEF.amova, nrepet = 999)
build_results_table(LpxA.MEF.amova)


PleD.MEF.amova = poppr.amova(PleD.MEF,hier=~Flow/Ind/Sample,clonecorrect = FALSE,within=F,threads=4,method='ade4')
# randtest(PleD.MEF.amova, nrepet = 999)
build_results_table(PleD.MEF.amova)

```


## AMOVA all regions

```{r}
LpxA.JdF.anova.region = poppr.amova(LpxA_obj,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
LpxA.JdF.anova.region
# randtest(LpxA.JdF.anova.region, nrepet = 999)
build_results_table(LpxA.JdF.anova.region)


PleD.JdF.anova.region = poppr.amova(PleD_obj,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
PleD.JdF.anova.region
# randtest(PleD.JdF.anova.region, nrepet = 999)
build_results_table(PleD.JdF.anova.region)


LpxA.JdF.anova.flow = poppr.amova(LpxA_obj,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
# LpxA.JdF.anova.flow
# randtest(LpxA.JdF.anova.flow, nrepet = 999)
build_results_table(LpxA.JdF.anova.flow)

PleD.JdF.anova.flow = poppr.amova(PleD_obj,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
# PleD.JdF.anova.flow
# randtest(PleD.JdF.anova.flow, nrepet = 999)
build_results_table(PleD.JdF.anova.flow)

```

## AMOVA witout MV
```{r}
MV = c('R15','R16')
END_samples = popNames(LpxA_obj)[!(popNames(LpxA_obj) %>% grepl(paste(MV, collapse="|"),.))]

LpxA_END = popsub(LpxA_obj, sublist=END_samples)
PleD_END = popsub(PleD_obj, sublist=END_samples)

LpxA.END.amova.region = poppr.amova(LpxA_END,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
# LpxA.END.amova.region
# randtest(LpxA.END.amova.region, nrepet = 999)
build_results_table(LpxA.END.amova.region)


LpxA.END.amova.flow = poppr.amova(LpxA_END,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
# LpxA.END.amova.flow
# randtest(LpxA.END.amova.flow, nrepet = 999)
build_results_table(LpxA.END.amova.flow)


PleD.END.amova.region = poppr.amova(PleD_END,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
# PleD.END.amova.region
# randtest(PleD.END.amova.region, nrepet = 999)
build_results_table(PleD.END.amova.region)


PleD.END.amova.flow = poppr.amova(PleD_END,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4')
# PleD.END.amova.flow
# randtest(PleD.END.amova.flow, nrepet = 999)
build_results_table(PleD.END.amova.flow)

```

# Plot ordinations

## PcoA on Phi-st distances (all samples)
```{r}
gene_phist= PleD_phist

dat=as.data.frame(pcoa(gene_phist)$vectors)[,1:2]
var_expl=pcoa(phist)$values$Relative_eig

dat['Sample']=rownames(dat)
dat=merge(dat,meta,by='Sample', all.x=T)
dat %>% head
dat$Region=factor(dat$Region,levels=c('MV','CB','MEF'))
dat$Flow=factor(dat$Flow,levels=c('H','L','B'))
dat$Habitat=paste(dat$Region,dat$Flow,sep='-')
dat$Habitat=factor(dat$Habitat,levels=c('MV-H','MV-L','CB-H','CB-B','MEF-H','MEF-L','MEF-B'))
  
habitat_colors=c(`MV-H` = "#a63603", 
`MV-L` = "#e6550d", `CB-H` = "#006d2c",`CB-B` = "#b2df8a",  
`MEF-H` = "#08519c", `MEF-L` = "#3182bd", `MEF-B` = "#9ecae1")

ggplot(dat)+
  geom_point(aes(x=Axis.1,y=Axis.2, color=Habitat, shape=Flow),size=3)+
  xlab(paste('PCoA 1 (',sprintf("%0.1f%%", var_expl[1]*100),')'))+
  ylab(paste('PCoA 2 (',sprintf("%0.1f%%", var_expl[2]*100),')'))+
  scale_color_manual(values = habitat_colors)+
  theme_bw()


```

## PcoA on Phi-st distances (witout MV)
```{r}
phist_nomv = LpxA_phist_nomv
phist_nomv = PleD_phist_nomv

dat=as.data.frame(pcoa(phist_nomv)$vectors)[,1:2]
str(pcoa(phist))
var_expl=pcoa(phist)$values$Relative_eig

dat['Sample']=rownames(dat)
dat=merge(dat,meta,by='Sample', all.x=T)
dat %>% head
dat$Region=factor(dat$Region,levels=c('MV','CB','MEF'))
dat$Flow=factor(dat$Flow,levels=c('H','L','B'))
dat$Habitat=paste(dat$Region,dat$Flow,sep='-')
dat$Habitat=factor(dat$Habitat,levels=c('MV-H','MV-L','CB-H','CB-B','MEF-H','MEF-L','MEF-B'))
  
habitat_colors=c(`MV-H` = "#a63603", 
`MV-L` = "#e6550d", `CB-H` = "#006d2c",`CB-B` = "#b2df8a",  
`MEF-H` = "#08519c", `MEF-L` = "#3182bd", `MEF-B` = "#9ecae1")

ggplot(dat)+
  geom_point(aes(x=Axis.1,y=Axis.2, color=Habitat, shape=Flow),size=3)+
  xlab(paste('PCoA 1 (',sprintf("%0.1f%%", var_expl[1]*100),')'))+
  ylab(paste('PCoA 2 (',sprintf("%0.1f%%", var_expl[2]*100),')'))+
  scale_color_manual(values = habitat_colors)+
  theme_bw()

```
