---
title: "CRISPR AMOVA analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Init environment and import packages
```{r setup, include=FALSE}
setwd("~/Desktop/CRISPR_distance/manuscript_data_and_downstream_analyses_scripts")
knitr::opts_chunk$set(echo = TRUE)
require('dplyr')
require('reshape2')
library('ggplot2')
require('ade4')
require('usedist') # package used to subset distance matrix
require('poppr')
require('vegan')
require('ape')
require('igraph')
```

# Init functions

```{r}
fix_phi_df <- function(df){
  df1 <- arrange(df, -row_number())
  rownames(df1) <- rev(rownames(df))
  df1=rbind(df1,data.frame(row.names = 'Total',Phi = NA))
  return(df1)
} # reorder rows of Phi-st output in poppr.amova

build_results_table <- function(res) {
  
  pval = randtest(res, nrepet =999)
  
  pval = rbind(data.frame(P.value=rev(pval$pvalue),row.names = rev(pval$names)),data.frame(row.names = 'Total',P.value = NA))
  
  table=cbind(res$results,res$componentsofcovariance,pval,fix_phi_df(res$statphi))
  
  return(table)
  
} # Parse results of poppr.amova

Get_pop_params = function(gid,palette='rainbow'){
  ## load functions we need from poppr
  .clonecorrector<-poppr:::.clonecorrector
  palette_parser<-poppr:::palette_parser
  
  ## Parse pop data
  mlgs <- mll(gid) 
  cgid   <- gid[.clonecorrector(gid), ]
  cmlg <- mll(cgid) #come up in same order as gid mlls
  npop   <- nPop(gid)
  npop   <- if (npop == 0) 1 else npop
  pnames <- popNames(gid)
  pnames <- if (is.null(pnames)) "pop" else pnames
    
  ## Generate color vector in palette name provided or use provided color vector, default in none provided = rainbow
    if(length(palette)==1){
        color  <- palette_parser(palette, npop, pnames)
    } else {color <- palette}
    mlg.number <- table(mlgs)[rank(cmlg)]
    names(mlg.number) <- paste0('MLG.', names(mlg.number))
    piece_of_pie <- !is.null(pnames) && npop > 1
    if (piece_of_pie){
      # Obtaining population information for all MLGs
      subs   <- sort(unique(mlgs))
      mlg.cp <- mlg.crosspop(gid, mlgsub = subs, quiet=TRUE)
      if (is.numeric(mlgs)){
        names(mlg.cp) <- paste0("MLG.", sort(unique(mlgs)))    
      }
      mlg.cp <- mlg.cp[rank(cmlg)]  
      # This creates a list of colors corresponding to populations.
      mlg.color <- lapply(mlg.cp, function(x) color[pnames %in% names(x)])
      # Set shape and circle color vectors
      mlg.sp        <- rep("pie", length(mlg.cp)) # set shape
      names(mlg.sp) <- names(mlg.cp)
      mlg.cc        <- rep(NA, length(mlg.cp)) # set circle color
      names(mlg.cc) <- names(mlg.cp)
      # Transform pie made of single population into circle
      pie.single    <- lengths(mlg.cp) == 1
      if (any(pie.single)) {
        mlg.sp[pie.single] <- "circle"
        mlg.cc[pie.single] <- unlist(mlg.color[pie.single])
      }
    }
  
  ## Get correspondance for MLG ids and haplo names
  haplo= strsplit(rownames(cgid@tab),'_') %>% sapply(tail, 2 ) %>% .[1,] %>% paste('',.,sep='') %>% as.numeric
  names(haplo) <- paste0('MLG.',cmlg)
  
  ## Build params object for igraph
  params=list('mlg.labels' = haplo,
              'mlg.sp' = mlg.sp,
              'mlg.cc' = mlg.cc,
              'mlg.number' = mlg.number,
              'mlg.color' = mlg.color,
              'mlg.cp' = mlg.cp)
  
  # mat = cbind(Haplo,mlg.sp ,mlg.cc, mlg.number) %>% as.data.frame
  return(params)
} # Get population parameters fron Genind object. Used for editing MST

```

# Import data. One haplotype per sequence read

```{r}
#### Import CRISPR Array distance matrix issued from Kupczok and Bollback (2013) algorithm ####

#### noneg v1 ####
path_to_CRISPR_arrays_distance_matrix = './CRISPRarrays_estimated_pairwiseDist.csv'
crispr_dist_all = read.csv(path_to_CRISPR_arrays_distance_matrix2,header=T,sep='\t',row.names=1)
colnames(crispr_dist_all)=c(1:length(colnames(crispr_dist_all))-1)
rownames(crispr_dist_all)=colnames(crispr_dist_all)
tail(crispr_dist_all)

crispr_dist_all<-as.dist(crispr_dist_all)
attr(crispr_dist_all,'Labels') 

#### Import CRISPR allele counts in populations ####
path_to_CRISPR_allele_counts_in_pop ='./ALLCRISPRs_haplo_matches_at_5_dist_V6_sorted.txt'

crispr_all = read.csv(path_to_CRISPR_allele_counts_in_pop,header=TRUE,sep=',')

colnames(crispr_all) = c('Array', colnames(crispr_all) %>% gsub('\\.','-',.) %>% .[2:length(.)])
crispr_all$`seq-id` <-NULL
crispr_all[is.na(crispr_all)] <- 0
rownames(crispr_all)=c(0:123)
crispr_all %>% tail
crispr_main = crispr_all[1:10,]
crispr_main

### Import metdata ###
path_to_CRISPR_metadata = './CRISPR_metadata.txt'

meta = read.csv(path_to_CRISPR_metadata,header=TRUE,sep='\t',as.is = T)
meta$Sample = gsub('^.{4}', '', meta$Sample)
meta$chunk = substr(meta$Sample, nchar(meta$Sample)-1, nchar(meta$Sample))
# meta$chunk[!(meta$dupl %in% c('s'))] = 'full'
meta$chunk[meta$chunk %in% c('-T','is','ld')] = 'full'
meta$Habitat <-paste(meta$Region,meta$Flow, sep = '-')
meta %>% head
str(meta)

#### Import per-read allele counts in populations. #### 


path_to_CRISPR_per_read_allele_counts_in_pop = './ALL_CRISPRs_haplo_matches_at_5_dist_V6_sorted_per_read.txt'

crispr_all_per_read <- read.table(path_to_CRISPR_per_read_allele_counts_in_pop,header=T,row.names = 1,sep='\t')


#### Import Phi-st distance matrix for PcoA analyses ####

path_to_Phist_matrix = './CRISPR_pairwisePhist.txt'

phist=read.csv(path_to_Phist_matrix,header=T,row.names = 1,sep='\t')
colnames(phist) = c(colnames(phist) %>% gsub('\\.','-',.) %>% .[1:length(.)])
phist=phist[order(rownames(phist)), order(colnames(phist))]
phist=as.dist(as.matrix(phist))
str(phist)

MV = c('R15','R16')
END_samples = labels(phist)[!(labels(phist) %>% grepl(paste(MV, collapse="|"),.))]

phist_nomv=dist_subset(phist,idx = END_samples)

```


# Species accumulation curves
Did we sample all of the diversity of CRISPR spacers in each trophosome chunk?

Getting accumulation curves on all 44 trophosome samples, we see only a few samples were sufficiently sampled to reveal their full diversity:
R07B-1-Ta, R07B-1-Td, R07B-2-Tb, R16L-5-T, R16L-2-T
The symbiont CRISPR diversity has not sufficiently been covered with the samples we took.
```{r}
library(vegan)
raremax <- min(rowSums(t(crispr_all)))

tcrispr=t(crispr_all)
tcrispr= tcrispr[order(rownames(tcrispr)),,drop=FALSE] %>% .[-1,]

a <- rarecurve(tcrispr[c(1:10),], step=1, cex=0.5) 
b <- rarecurve(tcrispr[c(10:20),], step=1, cex=0.5)
c <- rarecurve(tcrispr[c(30:40),], step=1, cex=0.5)
d <- rarecurve(tcrispr[c(35:44),], step=1, cex=0.5)
# rarecurve(t(crispr[,c(39:45)]), step = 20, sample = raremax, label = T)
```

However, the 10 main symbionts were sufficiently sampled with the 44 trophosome samples
 
```{r}
# With the samples we took, we have sufficiently covered the main symbiont diversity
sp2<-specaccum(t(crispr_main[,c(2:45)]), method="random")
plot(sp2)
mods <- fitspecaccum(sp2, "arrh")
plot(mods, col="hotpink")
boxplot(sp2, col = "yellow", border = "blue", lty=1, cex=0.3, add= TRUE)

```


# Duplicates 
Is the distribution of CRISPR haplotypes the same between technical duplicates? 
R02H-8-T_bis: same with chisq, not the same with fisher test --> these will not be merged
R15H-2-T:same --> the samples will be merged
R15H-3-T:same --> the samples will be merged

```{r}
#### CRISPR ####
dupl1=crispr_all[,c("R02H-8-T_bis","R02H-8-T")]
dupl1_nozero=dupl1[rowSums(dupl1)>0,]
chisq.test(dupl1_nozero)
fisher.test(dupl1)
diversity(t(dupl1),'simpson')  %>% mean(.)
diversity(t(dupl1))  %>% mean(.)

dupl2=crispr_all[,c("R15H-2-T_bis","R15H-2-T")]
dupl2_nozero=dupl2[rowSums(dupl2)>0,]
chisq.test(dupl2_nozero)
fisher.test(dupl2)
diversity(t(dupl2),'simpson') %>% mean(.)
diversity(t(dupl2))  %>% mean(.)


dupl3=crispr_all[,c("R15H-3-T_old","R15H-3-T")]
dupl3_nozero=dupl3[rowSums(dupl3)>0,]
chisq.test(dupl3_nozero)
fisher.test(dupl3)
diversity(t(dupl3),'simpson')
diversity(t(dupl3))  %>% mean(.)


```


# Create genind object

Add info about identical technical replicates to meta data
```{r}

meta$DedupInd<-ifelse(meta$Sample == 'R02H-8-T_bis','R02H-8-T_bis',meta$Ind)
  
head(meta)
```


```{r}
df = crispr_all_per_read
df %>% head
df$haplo %>% unique() %>% sort()
meta %>% head

per_read_meta = merge(df[,c('Sample','read_id.1')],meta[,c('Sample','Site','Ind','Region','Flow','Habitat','dupl','chunk','DedupInd')], by='Sample', all.x=T)
rownames(per_read_meta) = per_read_meta$read_id.1
per_read_meta$read_id.1 <- NULL
per_read_meta %>% head

df1=as.data.frame(df[,'haplo'],row.names = row.names(df))
colnames(df1)[1]='haplo'
df1 %>% head
obj <- df2genind(df1, ploidy=1,pop=factor(per_read_meta$Sample), strata=per_read_meta)

popsub(obj)
as.genclone(obj)
```

# Minimum Spanning Trees (MSTs)
## By Individual hosts
```{r}
Ind_obj<- clonecorrect(obj, strata = ~Ind, combine = TRUE, keep = 1) 
haplos_vec = strsplit(rownames(Ind_obj@tab),'_') %>% sapply(tail, 2 ) %>% .[1,]
reads_dist = as.matrix(crispr_dist_all)[haplos_vec,haplos_vec] %>% as.dist()

MST <- poppr.msn(Ind_obj,reads_dist,gadj = 15, vertex.label = 'MLG') # get initial MST
g <- MST$graph

Ind_pop<-clonecorrect(obj, strata = ~Sample, combine = TRUE, keep = 1) 
Ind_pop@pop<- Ind_pop@strata$Ind
params_full = Get_pop_params(Ind_pop,palette=colors)

## Edit MST

colors = c('#b3e1cc','#ccebc6','#f4f093','#7fb0d1','#bd5a2e','#d9d9d9','#ef8b5f','#a7762c','#fffccc','#cab0d4','#e4f3c6','#f7ccab','#bdaed4','#cccccc','#f3b260','#7dc77d','#55af4b','#d95f35','#999999','#bdb8d9','#666666','#fdeb6e','#e7468b','#974ca1','#b3de69','#fdf0ae','#df3b34','#f0e2cc','#899cc7')
names(colors) = as.factor(levels(Ind_obj@pop))
# library(scales)
# show_col(colors)

params = Get_pop_params(Ind_obj,palette=colors)

mlg.size = params_full$mlg.number[V(g)$label]
mlg.sp = params$mlg.sp[V(g)$label]
mlg.cc = params$mlg.cc[V(g)$label]
mlg.labels = params$mlg.labels[V(g)$label]
mlg.color = params$mlg.color[V(g)$label]
mlg.cp = params_full$mlg.cp[V(g)$label]
                            
## Apply params to MST


table = cbind(mlg.labels,mlg.size,mlg.cc,mlg.sp) %>% as.data.frame
table$mlg.mlgs = rownames(table)
rownames(table) = table$mlg.labels


cols.num <- c('mlg.labels', 'mlg.size')
table[cols.num] <- sapply(table[cols.num],as.numeric)
table['mlg.size'][is.na(table['mlg.size'])] <- 0
table['mlg.sp'][is.na(table['mlg.sp'])] <-'circle'
table$mlg.labels.main <- table$mlg.labels
table$mlg.labels.main[!(table$mlg.labels.main %in% c(0,1,2,3,4,5,6,7,8,9))]<-NA

rownames(table)<-table$mlg.mlgs
table = table[V(g)$label,]

par(mar=c(0,0,0,0))

# set.seed(2)
# set.seed(9)
set.seed(14)
plot <- plot.igraph(
        g,
        vertex.size = (log10(table$mlg.size+0.5) *10) ,
        vertex.shape = table$mlg.sp,
        vertex.color = table$mlg.cc,
        vertex.label = table$mlg.labels.main,
        vertex.label.color="black",

        vertex.pie = mlg.cp[table$mlg.mlgs],
        vertex.pie.color = mlg.color[table$mlg.mlgs],
        
        edge.arrow.size = 0,
        edge.width = 2,
        edge.color='grey',

        # layout =  layout_as_tree(g,root = c('0'),circular = T)
        # layout = layout_as_tree(g,mode='all',root=c('0'),circular=F)
        layout =  layout_with_fr
        


)

```

## By Habitats
```{r}
Hab_obj<- clonecorrect(obj, strata = ~Habitat, combine = TRUE, keep = 1) 
haplos_vec = strsplit(rownames(Hab_obj@tab),'_') %>% sapply(tail, 2 ) %>% .[1,]
reads_dist = as.matrix(crispr_dist_all)[haplos_vec,haplos_vec] %>% as.dist()

# commented in order to get same tree layout as By Individuals
# MST <- poppr.msn(Hab_obj,reads_dist,gadj = 15, vertex.label = 'MLG') # get initial MST
# g <- MST$graph

Hab_pop<-obj
Hab_pop@pop<- Hab_pop@strata$Habitat
params_full = Get_pop_params(Hab_pop,palette=colors)

## Edit MST

colors=c('#347fbd','#184f9c','#9ccae1','#326c2c','#b0de8a','#a63625','#e65537')
names(colors) = levels(Hab_obj$pop)

params = Get_pop_params(Hab_obj,palette=colors)
mlg.size = params_full$mlg.number[V(g)$label]
mlg.sp = params$mlg.sp[V(g)$label]
mlg.cc = params$mlg.cc[V(g)$label]
mlg.labels = params$mlg.labels[V(g)$label]
mlg.color = params$mlg.color[V(g)$label]
mlg.cp = params_full$mlg.cp[V(g)$label]


## Apply params to MST


table = cbind(mlg.labels,mlg.size,mlg.cc,mlg.sp) %>% as.data.frame
table$mlg.mlgs = rownames(table)
rownames(table) = table$mlg.labels

cols.num <- c('mlg.labels', 'mlg.size')
table[cols.num] <- sapply(table[cols.num],as.numeric)
table['mlg.size'][is.na(table['mlg.size'])] <- 0
table['mlg.sp'][is.na(table['mlg.sp'])] <-'circle'
table$mlg.labels.main <- table$mlg.labels
table$mlg.labels.main[!(table$mlg.labels.main %in% c(0,1,2,3,4,5,6,7,8,9))]<-NA

# totab = crispr_all[,-1] %>% rowSums()
# relab = decostand(t(crispr_all[,-1]), method = "total") %>% colSums()
# counts = cbind(totab,relab) %>% as.data.frame()
# colnames(counts)<-c('totab','relab')
# table = merge(table,counts, by=0, all.x=T)
rownames(table)<-table$mlg.mlgs
table = table[V(g)$label,]

par(mar=c(0,0,0,0))


# set.seed(2)
# set.seed(9)
set.seed(14)
plt <- plot.igraph(
        g,
        vertex.size =(log10(table$mlg.size)*4.5),

        vertex.shape = table$mlg.sp,
        vertex.color = table$mlg.cc,
        vertex.label = table$mlg.labels.main,
        vertex.label.color="black",

        vertex.pie = mlg.cp[table$mlg.mlgs],
        vertex.pie.color = mlg.color[table$mlg.mlgs],
        
        edge.arrow.size = 0,
        edge.width = 2,
        edge.color='grey',

        # layout =  layout_as_tree(g,root = c('0'),circular = T)
        layout =  layout_with_fr

)

```


# Run ade4's AMOVA analysis using the Poppr wrapper
## AMOVA in the MEF (sites within Flow + trophosome sections within hosts)
```{r}
CB_MV=c('R13','R14','R15','R16')

MEF_samples = popNames(obj)[!(popNames(obj) %>% grepl(paste(CB_MV, collapse="|"),.))]

MEF = popsub(obj, sublist=MEF_samples)

MEF.amova = poppr.amova(MEF,hier=~Flow/Site/Ind/chunk/Sample,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=dist_subset(crispr_dist_all,idx = MEF@all.names$haplo),squared=F,correction='lingoes')

# MEF.amova
# randtest(MEF.amova, nrepet = 999)
mefamova = build_results_table(MEF.amova)
mefamova
# ## converting a genind as data.frame
# genind2df(obj)
# genind2df(obj, sep="/")

```

## AMOVA all regions

```{r}
JdF.anova.region = poppr.amova(obj,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=crispr_dist_all, squared=F,correction='lingoes')
# JdF.anova.region
# randtest(JdF.anova.region, nrepet = 999)
a<- build_results_table(JdF.anova.region)


JdF.anova.flow = poppr.amova(obj,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=crispr_dist_all, squared=F,,correction='lingoes')
# JdF.anova.flow
# randtest(JdF.anova.flow, nrepet = 999)
b<-build_results_table(JdF.anova.flow)

```

## AMOVA witout MV
```{r}
MV = c('R15','R16')
END_samples = popNames(obj)[!(popNames(obj) %>% grepl(paste(MV, collapse="|"),.))]

END = popsub(obj, sublist=END_samples)

END.amova.region = poppr.amova(END,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=dist_subset(crispr_dist_all,idx = END@all.names$haplo),squared=F,correction='lingoes')
# END.amova.region
# randtest(END.amova.region, nrepet = 999)
c<-build_results_table(END.amova.region)


END.amova.flow = poppr.amova(END,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=dist_subset(crispr_dist_all,idx = END@all.names$haplo),squared=F,correction='lingoes')
# END.amova.flow
# randtest(END.amova.flow, nrepet = 999)
d<-build_results_table(END.amova.flow)

```


## AMOVA witout MV and LF samples; only HF vs B
```{r}
exclude = c('R15H','R16L', 'R01L', 'R09L')
END_samples = popNames(obj)[!(popNames(obj) %>% grepl(paste(exclude, collapse="|"),.))]

END = popsub(obj, sublist=END_samples)

END.amova.region = poppr.amova(END,hier=~Region/Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=dist_subset(crispr_dist_all,idx = END@all.names$haplo),squared=F,correction='lingoes')
# END.amova.region
# randtest(END.amova.region, nrepet = 999)
e<-build_results_table(END.amova.region)


END.amova.flow = poppr.amova(END,hier=~Flow/DedupInd,clonecorrect = FALSE,within=TRUE,threads=4,method='ade4',dist=dist_subset(crispr_dist_all,idx = END@all.names$haplo),squared=F,correction='lingoes')
# END.amova.flow
# randtest(END.amova.flow, nrepet = 999)
f<-build_results_table(END.amova.flow)

```


# Plot ordinations


## PcoA on Phi-st distances (all samples)
```{r}
dat=as.data.frame(pcoa(phist)$vectors)[,1:2]
str(pcoa(phist))
var_expl=pcoa(phist)$values$Relative_eig

dat['Sample']=rownames(dat)
dat=merge(dat,meta,by='Sample', all.x=T)
dat %>% head
dat$Region=factor(dat$Region,levels=c('MV','CB','MEF'))
dat$Flow=factor(dat$Flow,levels=c('H','L','B'))
dat$Habitat=paste(dat$Region,dat$Flow,sep='-')
dat$Habitat=factor(dat$Habitat,levels=c('MV-H','MV-L','CB-H','CB-B','MEF-H','MEF-L','MEF-B'))
  
habitat_colors=c(`MV-H` = "#a63603", 
`MV-L` = "#e6550d", `CB-H` = "#006d2c",`CB-B` = "#b2df8a",  
`MEF-H` = "#08519c", `MEF-L` = "#3182bd", `MEF-B` = "#9ecae1")

ggplot(dat)+
  geom_point(aes(x=Axis.1,y=Axis.2, color=Habitat, shape=Flow),size=3)+
  xlab(paste('PCoA 1 (',sprintf("%0.1f%%", var_expl[1]*100),')'))+
  ylab(paste('PCoA 2 (',sprintf("%0.1f%%", var_expl[2]*100),')'))+
  scale_color_manual(values = habitat_colors)+
  theme_bw()


```

## PcoA on Phi-st distances (witout MV)
```{r}
dat=as.data.frame(pcoa(phist_nomv)$vectors)[,1:2]
str(pcoa(phist))
var_expl=pcoa(phist)$values$Relative_eig

dat['Sample']=rownames(dat)
dat=merge(dat,meta,by='Sample', all.x=T)
dat %>% head
dat$Region=factor(dat$Region,levels=c('MV','CB','MEF'))
dat$Flow=factor(dat$Flow,levels=c('H','L','B'))
dat$Habitat=paste(dat$Region,dat$Flow,sep='-')
dat$Habitat=factor(dat$Habitat,levels=c('MV-H','MV-L','CB-H','CB-B','MEF-H','MEF-L','MEF-B'))
  
habitat_colors=c(`MV-H` = "#a63603", 
`MV-L` = "#e6550d", `CB-H` = "#006d2c",`CB-B` = "#b2df8a",  
`MEF-H` = "#08519c", `MEF-L` = "#3182bd", `MEF-B` = "#9ecae1")

ggplot(dat)+
  geom_point(aes(x=Axis.1,y=Axis.2, color=Habitat, shape=Flow),size=3)+
  xlab(paste('PCoA 1 (',sprintf("%0.1f%%", var_expl[1]*100),')'))+
  ylab(paste('PCoA 2 (',sprintf("%0.1f%%", var_expl[2]*100),')'))+
  scale_color_manual(values = habitat_colors)+
  theme_bw()

```

